Forecasting the United States 2024 Presidential Election Using State as Fixed Effect Through a Bayesain Approach

### 2024-11-04 paper/paper.qmd
fixed typos

---
title: "Forecasting the United States 2024 Presidential Election"
subtitle: "Kamala Harris wins 60th presidency with 279 Electoral College Votes"
author: 
  - Robert Ford
  - Michelle Ji
  - Cher Ning-Li
thanks: "Code and data are available at: https://github.com/Ford-Robert/us-presidential-election."
date: today
date-format: long
abstract: "This report analyzes compiled polling surveys for the 2024 U.S. presidential election using FiveThirtyEight data [@538data] and the 'pool of polls' method. From the analysis, bayesian and generalized linear models were built to forecast the outcome of the presidential election based on popular vote and the electoral college system. The models forecast Kamala Harris as the winner of election, winning 279 electoral college votes versus the 258 votes Donald Trump acquires. Election forecasting is crucial for informing voters, guiding campaign strategies, and engaging the public in the electoral process. Additionally, it helps predict potential policy shifts and holds political institutions accountable, contributing to a more transparent and informed democracy."
format:
  pdf:
    prefer-html: true
  html: default
header-includes:
  - \usepackage{geometry}
  - \geometry{margin=0.75in} 
number-sections: true
bibliography: references.bib
---

@@

Every four years, the United States presidential election is one of the most significant international political events. The complex election process involves both the popular and electoral college system. The candidate that receives the popular vote by the voters wins the electoral votes for that state [@electoralcollege]. After the popular vote is accounted for, the candidate who receives the majority of the electoral votes from all states wins the presidential election. The number of electoral votes varies by state and is proportional to each state’s respective population [@electoralcollege].

Specifically for the 2024 presidential election, the top candidates include Donald Trump and Kamala Harris, with differing opinions on key topics such as immigration, technology, abortion, and transatlantic affairs. For example, regarding technology, both parties are concerned with the accelerated progression of artificial intelligence’s capabilities but differ in the response and regulation of it. Republicans believe that the current practices are unfair and disproportionately target their voices, whereas Democrats generally are in favor of tighter regulation [@keyissues]. Another key issue important to voters and the candidates is abortion. As a former president, he supported restrictions on abortion access and advocated for policies that align with the conservative movement’s opposition to abortion rights [@abortionstance]. On the other hand, Harris has been a key supporter of the pro-choice movement and has been an advocate for protecting and expanding access to abortion services, as well as female autonomy [@harrisviews]. The outcome of this election will have profound consequences for various legislative policies and the future direction of the United States.

To forecast the 2024 United States presidential election, a generalized linear model and a Bayesian model were built. These models will help us predict who will win the election based on the electoral college system. An important finding from our model is that we predict Kamala Harris wins the presidential election with 279 electoral college votes. This would be a historic win, as she would be the first woman to become US president, a significant milestone in American politics. One drawback of our model stems from the representativeness of the dataset, as it does not include all 50 states in the polls, which may impact the electoral college votes. More is discussed in @sec-disc-weaknesses.

@@

The variables used in the dataset include pollster name, pollster weight, method, state, sample size, pollscore, collection/end date, transparency score, candidate, support percentage, election date, days to election, initial weight, weight, and poll region. Pollster Name indicates which pollster the individual poll came from and the weight is assigned by FiveThirtyEight depending on sample size and if they have multiple polls out in a short time to adjust for any bias. Method is the way the poll was deployed, including but not limited to live phone, online panel, and text-to-web. State indicates what state the poll was conducted and sample size is the size of the poll. The pollscore is the rating FiveThirtyEight gives each indidivual poll, on factors such on ethical and methodological elements.The candidate variable indicates the candidate a certain percentage of people in the poll voted for and "support percentage” represents that specific number. The election date and days to election variables indicate how far out the poll was conducted with respect with the election date of November 5, 2024. We also reorganized the dataset to include a poll region, which divides the US into geographical regions such as rust belt, northeast, etc. The table in @sec-app-table details the first few observations of the cleaned dataset, which we used to build our model.

Data was analyzed through the R programming software [@citeR] and packages such as `tidyverse` [@tidyverse], `ggplot2` [@ggplot2], `knitr` [@knitr], and `dplyr` [@dplyr] were used to help download, clean, simulate, analyze, and test the data. For model construction and analysis, the following packages were used: `rstanarm` [@rstanarm1; @rstanarm2], `brms` [@brms1; @brms2; @brms3], and `bayesplot` [@bayesplot1; @bayesplot2].

@@

All model construction and analysis was conducted using R [@citeR] with the following key packages:

- **tidyverse:** For data manipulation and visualization [@tidyverse].
- **rstanarm:** For Bayesian model fitting using Stan [@rstanarm1; @rstanarm2].
- **brms:** Alternative package for Bayesian regression modeling [@brms1; @brms2; @brms3].
- **bayesplot:** For extensive model diagnostics and visualization [@bayesplot1; @bayesplot2].

@@

For states lacking recent polling data, we incorporated historical averages of Democratic and Republican support from elections since 2000. This data was collected from a GitHub repository [@hist_state] which collected their data from Wikipedia. This decision is based on the assumption that certain states are unlikely to change their voting patterns, reducing the necessity for current polling data. The ten states that do not have polling data have not changed parties in at least the last 20-plus years (@sec-voting-hist). Alternatively, imputation methods or political regional averages could have been used, but historical averages are used for their simplicity and relevance.

@@

Ideally, multiple models would be created. Then, model selection techniques would be used to determine the best model for predicting the outcome of the US election. Due to time constraints, the following models were considered but not implemented. A set of Bayesian models with state as a random effect and models excluding specific predictors like Transparency Score. More complex models with additional interaction terms or non-linear components were also considered.

#### Assumptions and Limitations

@@

@fig-predic displays each candidate's chance of winning and the average number of Elector Votes they received over the 1000 simulated elections. Our model predicts that Harris has a 67.4% chance of winning the US presidential election. While Trump only has a 31.7% chance of winning. There is also a 0.9% chance of a tie On average across the 1,000 simulations Harris wins 279.33 Electoral Votes, while Trump wins on average 258.67 votes.

```{r}
#| label: fig-predic
#| fig-cap: Election Prediction Results
#| echo: false
#| warning: false
#| message: false

@@

Prediction betting markets provide a new data source that may be valuable for predicting the outcome of the US election. These online betting markets allow people to buy shares that represent an outcome of some event. If the event occurs then the share is worth $1, if not then the share's price falls to 0. This means the live share price reflects the probability of the event occurring. For example, for the outcome of the presidential election market, if Trump's share price is 65 cents this means the market is predicting there is a 65% chance that Trump will win the election. Recently the trading volume on these markets is significant, which indicates that these markets may be helpful in predicting election outcomes. Some papers have included data from these markets in their models to improve their prediction accuracy [@market_paper].

@@

## Deep Red/Blue State Voting History {#sec-voting-hist}
```{r}
#| fig-cap: Number of Wins in 1000 Simulations by State
#| echo: false
#| warning: false
#| message: false

#| fig-cap: Voting History of the 10 States That do not Have Poll Data
#| echo: false
#| warning: false
#| message: false

library(kableExtra)
library(dplyr)

# Create a data frame with the provided data
df <- data.frame(
  State = c("Alabama", "Delaware", "Hawaii", "Idaho", "Illinois",
            "Kentucky", "Louisiana", "Mississippi", "Tennessee", "Wyoming"),
  Party = c("Rep", "Dem", "Dem", "Rep", "Dem",
            "Rep", "Rep", "Rep", "Rep", "Rep"),
  Start_Year = c(1980, 1992, 1998, 1976, 1992,
                 2000, 2000, 1980, 2000, 1976),
  stringsAsFactors = FALSE
)

@@

# Combine into a single data frame with 6 columns
compact_df <- data.frame(
  State_1 = df1$State,
  Year_1 = df1$Start_Year,
  Party_1 = df1$Party,
  State_2 = df2$State,
  Year_2 = df2$Start_Year,
  Party_2 = df2$Party,
  stringsAsFactors = FALSE
)

# Function to assign color names based on party
assign_color <- function(party) {
  ifelse(party == "Rep", "#DD1717", "#0F4392")
}

# Apply cell_spec to each cell that needs coloring
compact_df <- compact_df %>%
  mutate(
    State_1 = cell_spec(State_1, background = assign_color(Party_1), color = "white", format = "latex", escape = FALSE),
    Year_1 = cell_spec(Year_1, background = assign_color(Party_1), color = "white", format = "latex", escape = FALSE),
    State_2 = cell_spec(State_2, background = assign_color(Party_2), color = "white", format = "latex", escape = FALSE),
    Year_2 = cell_spec(Year_2, background = assign_color(Party_2), color = "white", format = "latex", escape = FALSE)
  )

# Create and style the table
styled_table <- compact_df %>%
  select(State_1, Year_1, State_2, Year_2) %>%
  kable(format = "latex", booktabs = TRUE, escape = FALSE, align = "c",
        col.names = c("State", "Last Time Voted for a Different Party",
                      "State", "Last Time Voted for a Different Party")) %>%
  kable_styling(latex_options = c("striped"), full_width = FALSE, position = "left") %>%
  row_spec(0, bold = TRUE) %>%
  footnote(general = " Data collected only went back to 1976, some states may have been voting for the same party farther back than 1976.")

# Display the table
styled_table

```
## State By State Win {#sec-state-wins}

>>>>

---
title: "Forecasting the United States 2024 Presidential Election Using State as Fixed Effect Through a Bayesain Approach"
subtitle: "Kamala Harris wins 60th presidency with 279 Electoral College Votes"
author: 
  - Robert Ford
  - Michelle Ji
  - Cher Ning-Li
thanks: "Code and data are available at: https://github.com/Ford-Robert/us-presidential-election."
date: today
date-format: long
abstract: "This report analyzes compiled polling surveys for the 2024 U.S. presidential election using FiveThirtyEight data [@538data] and the 'pool of polls' method. From the analysis, bayesian and generalized linear models were built to forecast the outcome of the presidential election based on popular vote and the electoral college system. The models forecast Kamala Harris as the winner of election, winning 279 electoral college votes versus the 258 votes Donald Trump acquires. Election forecasting is crucial for informing voters, guiding campaign strategies, and engaging the public in the electoral process. Additionally, it helps predict potential policy shifts and holds political institutions accountable, contributing to a more transparent and informed democracy."
format:
  pdf:
    prefer-html: true
  html: default
header-includes:
  - \usepackage{geometry}
  - \geometry{margin=0.75in} 
number-sections: true
bibliography: references.bib
---

@@

Every four years, the United States presidential election is one of the most significant international political events. The complex election process involves both the popular and electoral college system. The candidate that receives the popular vote by the voters wins the electoral votes for that state [@electoralcollege]. After the popular vote is accounted for, the candidate who receives the majority of the electoral votes wins the presidential election. The number of electoral votes is proportional to each state’s respective population [@electoralcollege].

Specifically for the 2024 presidential election, the top candidates include Donald Trump and Kamala Harris, with differing opinions on key topics such as immigration, technology, abortion, and transatlantic affairs. For example, regarding technology, both parties are concerned with the accelerated progression of artificial intelligence’s capabilities but differ in the response and regulation of it. Republicans believe that current moderation practices are unfair and disproportionately target their voices, whereas Democrats generally are in favor of tighter regulation [@keyissues]. Another key issue important to voters and the candidates is abortion. As a former president, he supported restrictions on abortion access and advocated for policies that align with the conservative movement’s opposition to abortion rights [@abortionstance]. On the other hand, Harris has been a key supporter of the pro-choice movement and has been an advocate for protecting and expanding access to abortion services, as well as female autonomy [@harrisviews]. The outcome of this election will have profound consequences for various legislative policies and the future direction of the United States.

To forecast the 2024 United States presidential election a Bayesian model was built. These models will help us predict who will win the election based on the electoral college system. An important finding from our model is that we predict Kamala Harris wins the presidential election with 279 electoral college votes. This would be a historic win, as she would be the first woman to become US president, a significant milestone in American politics. One drawback of our model stems from the representativeness of the dataset, as it does not include all 50 states in the polls (after the cleaning process), which may impact the our estimates of electoral college votes. More is discussed in @sec-disc-weaknesses.

@@

The variables used in the dataset include pollster name, pollster weight, method, state, sample size, Poll Score, collection/end date, transparency score, candidate, support percentage, election date, days to election, initial weight, weight, and poll region. Pollster Name indicates which pollster the individual poll came from and the weight is assigned by FiveThirtyEight depending on sample size and if they have multiple polls out in a short time to adjust for any bias. Method is the way the poll was deployed, including but not limited to live phone, online panel, and text-to-web. State indicates what state the poll was conducted and sample size is the size of the poll. The Poll Score is the rating FiveThirtyEight gives each indidivual poll, on factors such on ethical and methodological elements.The candidate variable indicates the candidate a certain percentage of people in the poll voted for and "support percentage” represents that specific number. The election date and days to election variables indicate how far out the poll was conducted with respect with the election date of November 5, 2024. We also reorganized the dataset to include a poll region, which divides the US into geographical regions such as rust belt, northeast, etc. The table in @sec-app-table details the first few observations of the cleaned dataset, which we used to build our model.

Data was analyzed through the R programming software [@citeR] and packages such as tidyverse [@tidyverse], ggplot2 [@ggplot2], knitr [@knitr], and dplyr [@dplyr] were used to help download, clean, simulate, analyze, and test the data. For model construction and analysis, the following packages were used: rstanarm [@rstanarm1, @rstanarm2], brms [@brms1, @brms2, @brms3], and bayesplot [@bayesplot1, @bayesplot2]

@@

All model construction and analysis was conducted using R (@citeR) with the following key packages:

- **tidyverse:** For data manipulation and visualization. (@tidyverse)
- **rstanarm:** For Bayesian model fitting using Stan. (@rstanarm1, @rstanarm2)
- **brms:** Alternative package for Bayesian regression modeling. (@brms1, @brms2, @brms3)
- **bayesplot:** For extensive model diagnostics and visualization.(@bayesplot1, @bayesplot2)

@@

For states lacking recent polling data, we incorporated historical averages of Democratic and Republican support from elections since 2000. This data was collected from a GitHub repository (@hist_state) which collected their data from Wikipedia. This decision is based on the assumption that certain states are unlikely to change their voting patterns, reducing the necessity for current polling data. The ten states that do not have polling data have not changed parties in at least the last 20-plus years (@sec-voting-hist). Alternatively, imputation methods or political regional averages could have been used, but historical averages are used for their simplicity and relevance.

@@

Ideally, multiple models would be created. Then, model selection techniques would be used to determine the best model for predicting the outcome of the US election. Due to time constraints, the following models were considered but not implemented. A set of Bayesian models with state as a random effect and models excluding specific predictors like Transparency Score. More complex models with additional interaction terms or non-linear components were also considered

**Assumptions and Limitations**

@@

This table displays each candidate's chance of winning and the average number of Elector Votes they received over the 1000 simulated elections. Our model predicts that Harris has a 67.4% chance of winning the US presidential election. While Trump only has a 31.7% chance of winning. There is also a 0.9% chance of a tie On average across the 1,000 simulations Harris wins 279.33 Electoral Votes, while Trump wins on average 258.67 votes.

```{r}
#| label: table-2
#| fig-cap: Election Prediction Results
#| echo: false
#| warning: false
#| message: false

@@

Prediction betting markets provide a new data source that may be valuable for predicting the outcome of the US election. These online betting markets allow people to buy shares that represent an outcome of some event. If the event occurs then the share is worth $1, if not then the share's price falls to 0. This means the live share price reflects the probability of the event occurring. For example, for the outcome of the presidential election market, if Trump's share price is 65 cents this means the market is predicting there is a 65% chance that Trump will win the election. Recently the trading volume on these markets is significant, which indicates that these markets may be helpful in predicting election outcomes. Some papers have included data from these markets in their models to improve their prediction accuracy (@market_paper).

@@

## Deep Red/Blue State Voting History {#sec-voting-hist}
```{r}
#| fig-cap: Voting History of the 10 States That do not Have Poll Data
#| echo: false
#| warning: false
#| message: false

# Create the data frame
df <- data.frame(
  State = c("Alabama", "Delaware", "Hawaii", "Idaho", "Illinois",
            "Kentucky", "Louisiana", "Mississippi", "Tennessee", "Wyoming"),
  Party = c("Rep", "Dem", "Dem", "Rep", "Dem",
            "Rep", "Rep", "Rep", "Rep", "Rep"),
  Start_Year = c(1980, 1992, 1998, 1976, 1992,
                 2000, 2000, 1980, 2000, 1976),
  stringsAsFactors = FALSE
)

@@

# Combine into a single data frame with 4 columns
compact_df <- data.frame(
  State_1 = df1$State,
  Year_1 = df1$Start_Year,
  State_2 = df2$State,
  Year_2 = df2$Start_Year,
  stringsAsFactors = FALSE
)

# Load the knitr package
library(knitr)

# Create the table
kable(compact_df, col.names = c("State", "Last Time Voted for a Different Party",
                                "State", "Last Time Voted for a Different Party"))
```
## State By State Win {#sec-state-wins}
